---
title: 'R4DS: Chapter 4'
author: "Brandon Foltz"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

```{r}
flights|>
  filter(dep_delay > 120)
```

```{r}
flights|>
  filter(month == 1 & day == 1)
```

```{r}
flights |>
  filter(month == 1 | month == 2)
```

```{r}
flights|>
  filter(month %in% c(7,8))
```

```{r}
jan1 <- flights|>
  filter(month == 1)
```

```{r}
flights |>
  arrange(year, month, day, dep_time)
```

```{r}
flights|>
  arrange(desc(dep_delay))|>
  head()
```

```{r}
flights|>
  distinct()
```

```{r}
flights|>
  distinct(origin, dest)
```


```{r}
flights|>
  filter(dest == "CVG")
```

```{r}
flights|>
  filter(arr_delay > 120)|>
  filter(dest %in% c("IAH", "HOU"))|>
  filter(carrier %in% c("UA","DL","AA"))|>
  filter(month %in% c(7,8,9))|>
  head()
```

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )
```

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1 #before position / column 1
  )
```
```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .after = day
  )
```
```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used" # keeps just the vars in the mutate
  )
```

```{r}
flights |>
  select(month, day, year)
```
```{r}
flights|>
  select(day:year)
```
```{r}
flights|>
  select(!year:day)
```
```{r}
flights|>
  select(where(is.character))
```
There are a number of helper functions you can use within select():

starts_with("abc"): matches names that begin with “abc”.
ends_with("xyz"): matches names that end with “xyz”.
contains("ijk"): matches names that contain “ijk”.
num_range("x", 1:3): matches x1, x2 and x3.

Renaming a variable

```{r}
flights|>
  select(tail_num = tailnum)
```
```{r}
flights|>
  rename(tail_num = tailnum)
```
```{r}
flights|>
  relocate(time_hour, air_time)
```
```{r}
flights|>
  relocate(year:dep_time, .after = time_hour)
```
```{r}
flights|>
  relocate(starts_with("arr"), .before = dep_time)
```
```{r}
flights|>
  select(contains("TIME"))
```
4.4 The pip

Find the fast flights to Houston's IAH airport.

```{r}
flights|>
  filter(dest == "IAH")|>
  mutate(speed = distance / air_time * 60)|>
  select(year:day, dep_time, carrier, flight, speed)|>
  arrange(desc(speed))
```

Groups
group_by() changes the behavior of subsequent verbs
```{r}
flights|>
  group_by(month)
```
```{r}
flights|>
  group_by(month)|>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE)
  )
```
```{r}
flights|>
  group_by(month)|>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n()
  )
```
Slice

df |> slice_head(n = 1) takes the first row from each group.
df |> slice_tail(n = 1) takes the last row in each group.
df |> slice_min(x, n = 1) takes the row with the smallest value of column x.
df |> slice_max(x, n = 1) takes the row with the largest value of column x.
df |> slice_sample(n = 1) takes one random row.
Can also use prop = X.X to select a proportion

```{r}
flights|>
  group_by(dest)|>
  slice_max(arr_delay, n = 1)|>
  relocate(dest)
```
```{r}
daily <- flights|>
  group_by(year, month, day)
daily
```
```{r}
daily_flights <- daily|>
  summarise(n = n())
daily_flights
```
```{r}
daily|>
  ungroup()
```
```{r}
daily|>
  ungroup()|>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    flights = n()
  )
```
4.6 Case study: aggregates and sample size
https://r4ds.hadley.nz/data-transform.html#sec-sample-size

Specifically, we will compare what proportion of times a player gets a hit (H) vs. the number of times they try to put the ball in play (AB):

```{r}
batters <- Lahman::Batting|>
  group_by(playerID)|>
  summarise(
    performance = sum(H, na.rm = T) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )
batters
```
```{r}
batters|>
  filter(n > 100)|>
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1/10) +
  geom_smooth(se = FALSE)
```


