---
title: 'R4DS: Chapter 12'
author: "Brandon Foltz"
date: "2023-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(patchwork)
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency usually decreases with engine size")
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency usually decreases with engine size",
       subtitle = "Two seater sports cars are an exception due to their light weight",
       caption = "Data from fueleconomy.gov",
       x = "Engine displacement (L)",
       y = "Highway fuel economy (mpg)")
```

```{r}
df <- tibble(
  x = 1:10,
  y = x ^ 2
)

ggplot(df, aes(x, y)) +
  geom_point() +
  labs(
    x = quote(sum(x[i] ^ 2, i == 1, n)),
    y = quote(alpha + beta + frac(delta, theta))
  )
```

```{r}
ggplot(mpg,
       aes(x = cty, y = hwy, color = drv, shape = drv)) +
  geom_point() +
  labs(
    x = "City MPG",
    y = "Highway MPG"
  )
```

```{r}
label_info <- mpg|>
  group_by(drv)|>
  arrange(desc(displ))|>
  slice_head(n = 1)|>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front wheel drive",
      drv == "r" ~ "rear wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  )|>
  select(displ, hwy, drv, drive_type)

ggplot(mpg,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", vjust = "bottom"
  ) +
  theme(legend.position = "none")

ggplot(mpg,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", alpha = 0.5, nudge_y = 2,
  ) +
  theme(legend.position = "none")

ggplot(mpg,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, nudge_y = 2,
  ) +
  theme(legend.position = "none")
```

```{r}
potential_outliers <- mpg|>
  filter(hwy > 40 | (hwy > 20 & displ > 5))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text_repel(data = potential_outliers, aes(label = model)) +
  geom_point(data = potential_outliers, color = "red") +
  geom_point(data = potential_outliers, color = "red", size = 3, shape = "circle open")
```

```{r}
label_info <- mpg|>
  summarise(
    displ = max(displ),
    hwy = max(hwy),
    label = "Increasing engine size is \nrelated to decreasing fuel economy."
  )

ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text(
    data = label_info,
    aes(label = label),
    vjust = "top", hjust = "right"
  )
```
```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "text", x = Inf, y = Inf,
    label = "Increasing engine size is \nrelated to descreasing fuel economy.",
    vjust = "top", hjust = "right"
  )
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() +
  annotate(
    geom = "label", x = 4, y = 38,
    label = "Increasing engine size is \nrelated to decreasing fuel economy.
    \nUpward curve is due to high displacement, \nlight sports cars.",
    hjust = "left", color = "red"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "red",
    arrow = arrow(type = "closed")
  )
```

Use geom_hline() and geom_vline() to add reference lines. We often make them thick (linewidth = 2) and white (color = white), and draw them underneath the primary data layer. That makes them easy to see, without drawing attention away from the data.

Use geom_rect() to draw a rectangle around points of interest. The boundaries of the rectangle are defined by aesthetics xmin, xmax, ymin, ymax.

Use geom_segment() with the arrow argument to draw attention to a point with an arrow. Use aesthetics x and y to define the starting location, and xend and yend to define the end location.

12.4 Scales ----
https://r4ds.hadley.nz/communication.html#scales

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```

```{r}
ggplot(diamonds,
       aes(x = cut, y = price)) +
  geom_boxplot(alpha = 0.05) +
  scale_y_continuous(labels = scales::label_dollar())

ggplot(diamonds,
       aes(x = cut, y = price)) +
  geom_boxplot(alpha = 0.05) +
  scale_y_continuous(
    labels = scales::label_dollar(scale = 1/1000, suffix = "K"),
    breaks = seq(1000, 19000, 6000)
  )
```
```{r}
ggplot(diamonds,
       aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  scale_y_continuous(
    name = "Percentage",
    labels = scales::label_percent()
  )
```
```{r}
presidential|>
  mutate(id = 33 + row_number())|>
  ggplot(aes(x = start, y = id)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_x_date(name = NULL, breaks = presidential$start, date_labels = "'%y")
```

```{r}
base <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right")
base + theme(legend.position = "none")
```

```{r}
ggplot(mpg,
      aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 1, override.aes = list(size = 4)))
```

```{r}
ggplot(diamonds,
       aes(x = carat, y = price)) +
  geom_hex()

ggplot(diamonds,
       aes(x = log10(carat), y = log10(price))) +
  geom_hex()

ggplot(diamonds,
       aes(x = carat, y = price)) +
  geom_hex() +
  scale_x_log10() +
  scale_y_log10()
# this is best since it retains the original scale
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  scale_color_brewer(palette = "Set1")
```

```{r}
ggplot(mpg,
       aes(x =displ, y = hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_color_brewer(palette = "Set1")
```

```{r}
presidential|>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "red", Democratic = "blue"))
```

```{r}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)

ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed() +
  labs(title = "Default, continuous")

ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_c() +
  labs(title = "Viridis, continuous")

ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_b() +
  labs(title = "Viridis, binned")
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth() +
  coord_cartesian(xlim = c(5,7), ylim = c(10,30))

mpg|>
  filter(displ >=5, displ <= 7, hwy >= 10, hwy <= 30)|>
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth()
```

```{r}
suv <- mpg|>
  filter(class == "suv")
compact <- mpg|>
  filter(class == "compact")

ggplot(suv,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point()

ggplot(compact,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point()

x_scale <- scale_x_continuous(limits = range(mpg$displ))
y_scale <- scale_y_continuous(limits = range(mpg$hwy))
col_scale <- scale_color_discrete(limits = unique(mpg$drv))

ggplot(suv,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale

ggplot(compact,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale
```

12.5 Themes
https://r4ds.hadley.nz/communication.html#sec-themes

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  labs(
    title = "Highway mileage decreases as engine size increases",
    caption = "Source: https://fueleconomy.gov"
  ) +
  theme(
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal",
    legend.box.background = element_rect(colour = "black"),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0)
  )
```

```{r}
p1 <- ggplot(mpg,
             aes(x = displ, y = hwy)) +
  geom_point() +
  labs(title = "Plot 1")
p2 <- ggplot(mpg,
             aes(x = drv, y = hwy)) +
  geom_boxplot() +
  labs(title = "Plot 2")
p1 + p2

p3 <- ggplot(mpg,
             aes(x = cty, y = hwy)) +
  geom_point() +
  labs(title = "Plot 3")
(p1 | p3) / p2
```

```{r}
p1 <- ggplot(mpg, aes(x = drv, y = cty, color = drv)) + 
  geom_boxplot(show.legend = FALSE) + 
  labs(title = "Plot 1")

p2 <- ggplot(mpg, aes(x = drv, y = hwy, color = drv)) + 
  geom_boxplot(show.legend = FALSE) + 
  labs(title = "Plot 2")

p3 <- ggplot(mpg, aes(x = cty, color = drv, fill = drv)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Plot 3")

p4 <- ggplot(mpg, aes(x = hwy, color = drv, fill = drv)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Plot 4")

p5 <- ggplot(mpg, aes(x = cty, y = hwy, color = drv)) + 
  geom_point(show.legend = FALSE) + 
  facet_wrap(~drv) +
  labs(title = "Plot 5")

(guide_area() / (p1 + p2) / (p3 + p4) / p5) +
  plot_annotation(
    title = "City and highway mileage for cars with different drive trains",
    caption = "Source: https://fueleconomy.gov."
  ) +
  plot_layout(
    guides = "collect",
    heights = c(1, 3, 2, 4)
    ) &
  theme(legend.position = "top")
```

