---
title: 'R4DS: Chapter 14'
author: "Brandon Foltz"
date: "2023-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

Use parse_double() when you have numbers written as strings. Use parse_number() when the string contains non-numeric text.

```{r}
x  <- c("1.2", "5.6", "1e3")
parse_double(x)

x <- c("$1,234", "USD 3,513", "59%")
parse_number(x)
```

14.3 Counts
https://r4ds.hadley.nz/numbers.html#counts

```{r}
flights|>
  count(dest)

flights|>
  count(dest, sort = TRUE)
```
```{r}
flights|>
  group_by(dest)|>
  summarise(
    n = n(),
    delay = mean(arr_delay, na.rm = TRUE)
  )
```

```{r}
flights|>
  group_by(dest)|>
  summarise(carriers = n_distinct(carrier))|>
  arrange(desc(carriers))
```

A weighted count is a sum.

```{r}
flights|>
  group_by(tailnum)|>
  summarise(miles = sum(distance))|>
  arrange(desc(miles))

flights|>
  count(tailnum, wt = distance)|>
  arrange(desc(n))
```
```{r}
flights|>
  group_by(dest)|>
  summarise(n_cancelled = sum(is.na(dep_time)))
```

Smallest or largest value in each row; pmin() and pmax()

```{r}
df <- tribble(
  ~x, ~y,
  1,  3,
  5,  2,
  7, NA,
)

df |> 
  mutate(
    min = pmin(x, y, na.rm = TRUE),
    max = pmax(x, y, na.rm = TRUE)
  )
```

When you do it wrong

```{r}
df |> 
  mutate(
    min = min(x, y, na.rm = TRUE),
    max = max(x, y, na.rm = TRUE)
  )
```

Modular artimetic

```{r}
1:10 %/% 3
1:10 %% 3
```

```{r}
flights|>
  mutate(
    hour = sched_dep_time %/% 100,
    minute = sched_dep_time %% 100,
    .keep = "used"
  )
```

```{r}
flights|>
  group_by(hour = sched_dep_time %/% 100)|>
  summarise(prop_cancelled = mean(is.na(dep_time)), n = n())|>
  filter(hour > 1)|>
  ggplot(aes(x = hour, y = prop_cancelled)) +
  geom_line(color = "grey50")+
  geom_point(aes(size = n))
```
Rounding

```{r}
round(123.456, 2)  # two digits
#> [1] 123.46
round(123.456, 1)  # one digit
#> [1] 123.5
round(123.456, -1) # round to nearest ten
#> [1] 120
round(123.456, -2) # round to nearest hundred
#> [1] 100
```
```{r}
x <- 123.456

floor(x)
#> [1] 123
ceiling(x)
#> [1] 124

# Round down to nearest two digits
floor(x / 0.01) * 0.01
#> [1] 123.45
# Round up to nearest two digits
ceiling(x / 0.01) * 0.01
#> [1] 123.46

# Round to nearest multiple of 4
round(x / 4) * 4
#> [1] 124

# Round to nearest 0.25
round(x / 0.25) * 0.25
#> [1] 123.5
```

CUtting numbers into ranges

```{r cutting_numbers_ranges}
x <- c(1, 2, 5, 10, 15, 20)
cut(x, breaks = c(0, 5, 10, 15, 20))
cut(x, breaks = c(0, 5, 10, 100))

cut(x, 
  breaks = c(0, 5, 10, 15, 20), 
  labels = c("sm", "md", "lg", "xl")
)

y <- c(NA, -10, 5, 10, 30)
cut(y, breaks = c(0, 5, 10, 15, 20))
```

Cumulative sums and rolling aggregates

```{r}
x <- 1:10
x
cumsum(x)
cumprod(x)
cummin(x)
cummax(x)
cummean(x)

y = rnorm(100)*100
cummean(y)
```

General Transformations

```{r}
x <- c(1, 2, 2, 3, 4, NA)
min_rank(x)
min_rank(desc(x))

df <- tibble(x = x)
df |> 
  mutate(
    row_number = row_number(x),
    dense_rank = dense_rank(x),
    percent_rank = percent_rank(x),
    cume_dist = cume_dist(x)
  )

df <- tibble(id = 1:10)

df |> 
  mutate(
    row0 = row_number() - 1,
    three_groups = row0 %% 3,
    three_in_each_group = row0 %/% 3
  )
```

Offsets: lag() and lead() and differencing

```{r}
x <- c(2, 5, 11, 11, 19, 35)
lag(x)
lead(x)
x - lag(x) # differencing
x == lag(x) # value changes along the vector
```

Consecutive Identifiers

```{r}
events <- tibble(
  time = c(0, 1, 2, 3, 5, 10, 12, 15, 17, 19, 20, 27, 28, 30)
)

# detect gaps of 5 or greater along the vector

events <- events |> 
  mutate(
    diff = time - lag(time, default = first(time)),
    has_gap = diff >= 5
  )
events

events |> mutate(
  group = cumsum(has_gap)
)

#counts, ie.increments group by 1 if there is a gap
```

```{r}
# create grouping variables

df <- tibble(
  x = c("a", "a", "a", "b", "c", "c", "d", "e", "a", "a", "b", "b"),
  y = c(1, 2, 3, 2, 4, 1, 3, 9, 4, 8, 10, 199)
)

df |> 
  group_by(id = consecutive_id(x)) |> 
  slice_head(n = 1)
```

Numeric Summaries

```{r}
flights|>
  group_by(year, month, day)|>
  summarise(
    mean = mean(dep_delay, na.rm = TRUE),
    median = median(dep_delay, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )|>
  ggplot(
    aes(x = mean, y = median)) +
      geom_abline(slope = 1, intercept = 0, color = "white", linewidth = 2) +
      geom_point()
```

Min, max, and quantiles

```{r}
flights|>
  group_by(year, month, day)|>
  summarise(
    max = max(dep_delay, na.rm = TRUE),
    min = min(dep_delay, na.rm = TRUE),
    q95 = quantile(dep_delay, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```

Spread

```{r}
flights|>
  group_by(origin, dest)|>
  summarise(
    distance_IQR = IQR(distance),
    n = n(),
    .groups = "drop"
  )|>
  filter(distance_IQR > 0)
```

Distributions

```{r}
# subgroups for each day should be similar

flights|>
  filter(dep_delay < 120)|>
  ggplot(
    aes(x = dep_delay, group = interaction(day, month)))+
  geom_freqpoly(binwidth = 5, alpha = 1/5)
```

Positions

```{r}
flights|>
  group_by(year, month, day)|>
  summarise(
    first_dep = first(dep_time, na_rm = TRUE),
    fifth_dep = nth(dep_time, 5, na_rm = TRUE),
    last_dep = last(dep_time, na_rm = TRUE)
  )
```

```{r}
flights|>
  group_by(year, month, day)|>
  mutate(r = min_rank(desc(sched_dep_time)))|>
  filter(r %in% c(1, max(r)))
```

With mutate()

x / sum(x) calculates the proportion of a total.
(x - mean(x)) / sd(x) computes a Z-score (standardized to mean 0 and sd 1).
(x - min(x)) / (max(x) - min(x)) standardizes to range [0, 1].
x / first(x) computes an index based on the first observation.

