---
title: 'R4DS: Chapter 16'
author: "Brandon Foltz"
date: "2023-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(babynames)
```

```{r}
str_view(fruit, "berry")
str_view(c("a", "ab", "ae", "bd", "ea", "eab"), "a.")
str_view(fruit, "a...e")
```

? makes a pattern optional (i.e. it matches 0 or 1 times)
+ lets a pattern repeat (i.e. it matches at least once)
* lets a pattern be optional or repeat (i.e. it matches any number of times, including 0).

```{r quantifiers}
# ab? matches an "a", optionally followed by a "b".
str_view(c("a", "ab", "abb"), "ab?")

# ab? matches an "a", optionally followed by a "b".
str_view(c("a", "ab", "abb"), "ab+")

# ab? matches an "a", optionally followed by a "b".
str_view(c("a", "ab", "abb"), "ab*")
```
Key Functions

```{r}
str_detect(c("a","b","c"), "[aeiou]")

babynames|>
  filter(str_detect(name, "x"))|>
  count(name, wt = n, sort = TRUE)

babynames|>
  group_by(year)|>
  summarise(prop_x = mean(str_detect(name, "x")))|>
  ggplot(aes(
    x = year,
    y = prop_x
  )) +
  geom_line() +
  labs(
    title = "Baby names containing an 'x' have increased substantially",
    x = "Year",
    y = "Proportion of baby names"
  )

str_subset(c("a","b","c"), "[aeiou]")
str_which(c("a", "b", "c"), "[aeiou]")
```
Count Matches

```{r}
x <- c("apple", "banana", "pear")
str_count(x, "p")

str_count("abababa", "aba")
str_view("abababa", "aba")

babynames|>
  count(name)|>
  mutate(
    vowels = str_count(name, "[aeiouAEIOU]"),
    consonants = str_count(name, "[^aeiouAEIOU]")
  )

babynames|>
  count(name)|>
  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]")
  )
```
Replace values

```{r}
x <- c("apple", "pear", "banana")
str_replace_all(x, "[aeiou]", "-")

str_remove_all(x, "[aeiou]")
```

Extract Variables

```{r}
df <- tribble(
  ~str,
  "<Sheryl>-F_34",
  "<Kisha>-F_45", 
  "<Brandon>-N_33",
  "<Sharon>-F_38", 
  "<Penny>-F_58",
  "<Justin>-M_41", 
  "<Patricia>-F_84", 
)

df|>
  separate_wider_regex(
    str,
    patterns = c(
      "<",
      name = "[A-Za-z]+",
      ">-",
      gender = ".", "_",
      age = "[0-9]+"
    )
  )
```

16.4 Pattern Details

```{r}
dot <- "\\."
str_view(dot)
str_view(c("abc", "a.c", "bef"), "a\\.c")

x <- "a\\b"
str_view(x)
str_view(x, "\\\\")
str_view(x, r"{\\}")

str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")
str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
```
Anchors

```{r}
str_view(fruit, "^a")
str_view(fruit, "a$")

str_view(fruit, "apple")
str_view(fruit, "^apple$")

x <- c("summary(x)", "summarize(df)", "rowsum(x)", "sum(x)")
str_view(x, "sum")

str_view(x, "\\bsum\\b")

str_view("abc", c("$", "^", "\\b"))

str_replace_all("abc", c("$", "^", "\\b"), "--")

```
Character Classes

```{r}
x <- "abcd ABCD 12345 -!@#%."
str_view(x, "[abc]+")

str_view(x, "[a-z]+")

str_view(x, "[^a-z0-9]+")


# You need an escape to match characters that are otherwise
# special inside of []
str_view("a-b-c", "[a-c]")

str_view("a-b-c", "[a\\-c]")

```
\d matches any digit;
\D matches anything that isn’t a digit.
\s matches any whitespace (e.g. space, tab, newline);
\S matches anything that isn’t whitespace.
\w matches any “word” character, i.e. letters and numbers;
\W matches any “non-word” character.

```{r}
x <- "abcd ABCD 12345 -!@#%."
str_view(x, "\\d+")

str_view(x, "\\D+")

str_view(x, "\\s+")

str_view(x, "\\S+")

str_view(x, "\\w+")

str_view(x, "\\W+")

```
Quantifiers

{n} matches exactly n times.
{n,} matches at least n times.
{n,m} matches between n and m times.

```{r}
x <- "-- -x- -xx- -xxx- -xxxx- -xxxxx-"
str_view(x, "-x?-")      
str_view(x, "-x+-")     
str_view(x, "-x*-")    
str_view(x, "-x{2}-")    
str_view(x, "-x{2,}-")  
str_view(x, "-x{2,3}-")
```
Grouping and capturing

```{r}
str_view(fruit, "(..)\\1")
str_view(words, "^(..).*\\1$")

#switch 2nd and 3rd words
sentences|>
  str_replace("(\\w+) (\\w+) (\\w+)", "\\1 \\3 \\2")|>
  str_view()

sentences|>
  str_match("the (\\w+) (\\w+)")|>
  head()

sentences|>
  str_match("the (\\w+) (\\w+)")|>
  as_tibble(.name_repair = "minimal") |>
  set_names("match", "word1", "word2")

x <- c("a gray cat", "a grey dog")
str_match(x, "gr(e|a)y")

str_match(x, "gr(?:e|a)y")
```

16.5 Pattern Control

```{r}
bananas <- c("banana", "Banana", "BANANA")
str_view(bananas, "banana")
str_view(bananas, regex("banana", ignore_case = TRUE))

x <- "Line 1\nLine 2\nLine 3"
str_view(x, ".Line")
str_view(x, regex(".Line", dotall = TRUE))

x <- "Line 1\nLine 2\nLine 3"
str_view(x, "^Line")
str_view(x, regex("^Line", multiline = TRUE))
```

```{r}
phone <- regex(
  r"(
    \(?     # optional opening parens
    (\d{3}) # area code
    [)\-]?  # optional closing parens or dash
    \ ?     # optional space
    (\d{3}) # another three numbers
    [\ -]?  # optional space or dash
    (\d{4}) # four more numbers
  )",
  comments = TRUE
)

str_extract(c("514-791-8141", "(123) 456 7890", "123456"), phone)
```

```{r}
str_view(c("", "a", "."), fixed("."))

str_view("x X", "X")
str_view("x X", fixed("X", ignore_case = TRUE))
```

Practice

```{r}
str_view(sentences, "^The")
str_view(sentences, "^The\\b")
str_view(sentences, "^She|He|It|They\\b")
str_view(sentences, "^(She|He|It|They)\\b")
```
```{r}
str_view(words, "^[^aeiou]+$")
str_view(words[!str_detect(words, "[aeiou]")])

str_view(words, "a.*b|b.*a")
words[str_detect(words, "a") & str_detect(words, "b")]

words[
  str_detect(words, "a") &
  str_detect(words, "e") &
  str_detect(words, "i") &
  str_detect(words, "o") &
  str_detect(words, "u")
]
```

Creating a pattern with code

```{r}
str_view(sentences, "\\b(red|green|blue)\\b")
rgb <- c("red", "green", "blue")

str_c("\\b(", str_flatten(rgb, "|"), ")\\b")
str_view(colors())
cols <- colors()
cols <- cols[!str_detect(cols, "\\d")]
str_view(cols)

pattern <- str_c("\\b(", str_flatten(cols, "|"), ")\\b")
str_view(sentences, pattern)
```

