---
title: 'R4DS: Chapter 19'
author: "Brandon Foltz"
date: "2023-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

```{r}
treatment <- tribble(
  ~person, ~treatment, ~response,
  "Derrick Whitmore", 1, 7,
  NA, 2, 10,
  NA, 3, NA,
  "Katherine Burke", 1, 4
)
```
When data is entered by hand, missing values sometimes indicate that the value in the previous row has been repeated (or carried forward):

```{r}
treatment|>
  fill(everything())
```

```{r}
x <- c(1,4,5,7,NA)
coalesce(x, 0)

x <- c(1,4,5,7,-99)
na_if(x, -99)

```

This dataset has two missing observations:

The price in the fourth quarter of 2020 is explicitly missing, because its value is NA.

The price for the first quarter of 2021 is implicitly missing, because it simply does not appear in the dataset.

```{r}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

```{r}
stocks|>
  pivot_wider(
    names_from = qtr,
    values_from = price
  )
```

tidyr::complete() allows you to generate explicit missing values by providing a set of variables that define the combination of rows that should exist.
```{r}
stocks|>
  complete(year, qtr)

stocks|>
  complete(year = 2019:2021, qtr)
```

Anti-joins to find missing values

```{r}
# missing information for four airports
flights|>
  distinct(faa = dest)|>
  anti_join(airports)

# missing information for 722 planes
flights|>
  distinct(tailnum)|>
  anti_join(planes)
```

Factors and empty groups
```{r}
health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)

#keep empty groups

health|>
  count(smoker, .drop = FALSE)

ggplot(health,
       aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)

health|>
  group_by(smoker, .drop = FALSE)|>
  summarise(
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  )
```

```{r}
health|>
  group_by(smoker)|>
  summarise(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  )|>
  complete(smoker)
```

