---
title: 'R4DS: Chapter 20'
author: "Brandon Foltz"
date: "2023-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

Finding keys using a simple count and checking for NA's which will not show up in count.
```{r}
planes|>
  count(tailnum)|>
  filter(n > 1)

weather|>
  count(time_hour, origin)|>
  filter(n > 1)

planes|>
  filter(is.na(tailnum))

weather|>
  filter(is.na(time_hour) | is.na(origin))
```

Surrogate Keys

```{r}
flights|>
  count(time_hour, carrier, flight)|>
  filter(n > 1)

airports|>
  count(alt, lat)|>
  filter(n > 1)

#surrogate key that is the row number, like an index
flights2 <- flights|>
  mutate(id = row_number(), .before = 1)
flights2
```
Basic Joins

```{r}
flights2 <- flights|>
  select(year, time_hour, origin, dest, tailnum, carrier)
flights2

flights2|>
  left_join(airlines)

flights2|>
  left_join(weather|>
  select(origin, time_hour, temp, wind_speed))

flights2|>
  left_join(planes|>
              select(tailnum, type, engines, seats))

flights2|>
  filter(tailnum == "N3ALAA")|>
  left_join(planes|>
              select(tailnum, type, engines, seats))
```

Specifying Join Keys
```{r}
flights2|>
  left_join(planes)

flights2|>
  left_join(planes, join_by(tailnum))

flights2|>
  left_join(airports, join_by(dest == faa))

flights2|>
  left_join(airports, join_by(origin == faa))
```

Filtering Joins

```{r}
airports|>
  semi_join(flights2, join_by(faa==origin))

airports|>
  semi_join(flights2, join_by(faa==dest))

#For example, we can find rows that are missing from airports by looking for flights that don’t have a matching destination airport:

flights2|>
  anti_join(airports, join_by(dest==faa))|>
  distinct(dest)

# Or we can find tailnums that are missing from planes

flights2|>
  anti_join(planes, join_by(tailnum))|>
  distinct(tailnum)
```

How do joins work?

```{r}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```
Non-eqi joins

```{r}
x |> left_join(y, by = "key", keep = TRUE)

```

Cross Joins

```{r}
# cartesian product of a list (permutations)

df <- tibble(name = c("John", "Simon", "Tracy", "Max"))
df |> cross_join(df)

# inequality join (combinations)

df <- tibble(id = 1:4, name = c("John", "Simon", "Tracy", "Max"))

df |> left_join(df, join_by(id < id))
```

Rolling Join

For example, imagine that you’re in charge of the party planning commission for your office. Your company is rather cheap so instead of having individual parties, you only have a party once each quarter. The rules for determining when a party will be held are a little complex: parties are always on a Monday, you skip the first week of January since a lot of people are on holiday, and the first Monday of Q3 2022 is July 4, so that has to be pushed back a week. That leads to the following party days:

```{r}
parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03"))
)

employees <- tibble(
  name = sample(babynames::babynames$name, 100),
  birthday = ymd("2022-01-01") + (sample(365, 100, replace = TRUE) - 1)
)
employees

#And for each employee we want to find the first party date that comes after (or on) their birthday. We can express that with a rolling join:
  
employees |> 
  left_join(parties, join_by(closest(birthday >= party)))
```

Overlap Joins

```{r}
parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03")),
  start = ymd(c("2022-01-01", "2022-04-04", "2022-07-11", "2022-10-03")),
  end = ymd(c("2022-04-03", "2022-07-11", "2022-10-02", "2022-12-31"))
)
parties

parties|>
  inner_join(parties, join_by(overlaps(start, end, start, end), q < q))|>
  select(start.x, end.x, start.y, end.y)

parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03")),
  start = ymd(c("2022-01-01", "2022-04-04", "2022-07-11", "2022-10-03")),
  end = ymd(c("2022-04-03", "2022-07-10", "2022-10-02", "2022-12-31"))
)

employees|>
  inner_join(parties, join_by(between(birthday, start, end)), unmatched = "error")
```

