---
title: 'R4DS: Chapter 26'
author: "Brandon Foltz"
date: "2023-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

```{r}
df <- tibble(
  a = rnorm(5),
  b = rnorm(5),
  c = rnorm(5),
  d = rnorm(5),
)
```
 Repeated parts
 
(a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE))
(b - min(b, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE))
(c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE))
(d - min(d, na.rm = TRUE)) / (max(d, na.rm = TRUE) - min(d, na.rm = TRUE))  

(█ - min(█, na.rm = TRUE)) / (max(█, na.rm = TRUE) - min(█, na.rm = TRUE))

To turn this into a function you need three things:

A name. Here we’ll use rescale01 because this function rescales a vector to lie between 0 and 1.

The arguments. The arguments are things that vary across calls and our analysis above tells us that we have just one. We’ll call it x because this is the conventional name for a numeric vector.

The body. The body is the code that’s repeated across all the calls.

```{r}
name <- function(arguments){
  body
}
```

```{r}
rescale01 <- function(x){
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

rescale01(c(-10, 0 , 10))

rescale01(c(1,2,3,NA,5))
```

```{r}
df|>
  mutate(
    a = rescale01(a),
    b = rescale01(b),
    c = rescale01(c),
    d = rescale01(d)
  )
```
Improve it

```{r}
rescale01  <- function(x){
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

# the first element of rng is the min, the second is the max
```

```{r}
x <- c(1:10, Inf)
rescale01(x)
```
Improve it more

```{r}
rescale01 <- function(x){
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale01(x)
```
 
Mutate Functions

```{r}
z_score <- function(x){
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}
```
For example, this clamp() function ensures all values of a vector lie in between a minimum or a maximum:

```{r}
clamp <- function(x, min, max){
  case_when(
    x < min ~ max,
    x > max ~ min,
    .default = x
  )
}

clamp(1:10, min = 3, max = 7)
```

Maybe you need to make the first character upper case:

```{r}
first_upper <- function(x){
  str_sub(x, 1, 1) <- str_to_upper(str_sub(x,1,1))
  x
}

first_upper("hello")
```

Or maybe you want to strip percent signs, commas, and dollar signs from a string before converting it into a number:

```{r}
clean_number <- function(x){
  is_pct <- str_detect(x, "%") # detect a percentage boolean
  num <- x|>
    str_remove_all("%")|> 
    str_remove_all(",")|>
    str_remove_all(fixed("$"))|>
    as.numeric(x)
  if_else(is_pct, num / 100, num)
}

clean_number("$12,300")
clean_number("45%")
```

For example, if you have a bunch of variables that record missing values as 997, 998, or 999, you might want to write a function to replace them with NA:

```{r}
fix_na <- function(x){
  if_else(x %in% c(997, 998, 999), NA, x)
}
```

Summary functions

```{r}
commas <- function(x){
  str_flatten(x, collapse = ", ", last = " and ")
}

commas(c("cat", "dog", "pigeon"))
```

Or you might wrap up a simple computation, like for the coefficient of variation, which divides the standard deviation by the mean:

```{r}
cv <- function(x, na.rm = FALSE){
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}

cv(runif(100, min = 0, max = 50))
cv(runif(100, min = 0, max = 500))
```

Or maybe you just want to make a common pattern easier to remember by giving it a memorable name:

```{r}
n_missing <- function(x){
  sum(is.na(x))
}

n_missing(c(4,NA,8,2,NA,10))
```

You can also write functions with multiple vector inputs. For example, maybe you want to compute the mean absolute prediction error to help you compare model predictions with actual values:

```{r}
mape <- function(actual, predicted){
  sum(abs((actual - predicted) / actual)) / length(actual)
}

mape(c(13,16,15,18,20), c(15,10,21,13,16))
```

```{r}
grouped_mean <- function(df, group_var, mean_var){
  df|>
    group_by({{group_var}})|>
    summarise(mean({{mean_var}}))
}

df <- tibble(
  mean_var = 1,
  group_var = "g",
  group = 1,
  x = 10,
  y = 100
)

df|>
  grouped_mean(group, x)
```

```{r}
summary6 <- function(data, var){
  data|>
    summarise(
      min = min({{var}}, na.rm = TRUE),
      mean = mean({{var}}, na.rm = TRUE),
      median = median({{var}}, na.rm = TRUE),
      max = max({{var}}, na.rm = TRUE),
      n = n(),
      n_miss = sum(is.na({{var}})),
      .groups = "drop"
    )
}

diamonds|>
  summary6(carat)

diamonds|>
  group_by(cut)|>
  summary6(carat)

diamonds|>
  group_by(cut)|>
  summary6(log10(carat))
```

Another popular summarize() helper function is a version of count() that also computes proportions:

```{r}
count_prop <- function(df, var, sort = FALSE){
  df|>
    count({{var}}, sort = sort)|>
    mutate(prop = n / sum(n))
}

diamonds|>
  count_prop(clarity)
```

Or maybe you want to find the sorted unique values of a variable for a subset of the data. Rather than supplying a variable and a value to do the filtering, we’ll allow the user to supply a condition:

```{r}
unique_where <- function(df, condition, var){
  df|>
    filter({{condition}})|>
    distinct({{var}})|>
    arrange({{var}})
}

# Find all the destination in December

flights|>
  unique_where(month==12, dest)
```

```{r}
subset_flights <- function(rows, cols){
  flights|>
    filter({{rows}})|>
    select(time, hour, carrier, flight, {{cols}})
}
```

For example, imagine you want to write a count_missing() that counts the number of missing observations in rows. You might try writing something like:

```{r}
count_missing <- function(df, group_vars, x_var){
  df|>
    group_by({{group_vars}})|>
    summarise(
      n_miss = sum(is.na({{x_var}})),
      .groups = "drop"
    )
}

flights|>
  count_missing(c(year, month, day), dep_time)

count_missing <- function(df, group_vars, x_var){
  df|>
    group_by(pick({{group_vars}}))|>
    summarise(
      n_miss = sum(is.na({{x_var}})),
      .groups = "drop"
    )
}

flights|>
  count_missing(c(year, month, day), dep_time)
```

Another convenient use of pick() is to make a 2d table of counts. Here we count using all the variables in the rows and columns, then use pivot_wider() to rearrange the counts into a grid:

```{r}
count_wide <- function(data, rows, cols){
  data|>
    count(pick(c({{rows}}, {{cols}})))|>
    pivot_wider(
      names_from = {{cols}},
      values_from = n,
      names_sort = TRUE,
      values_fill = 0
    )
}

diamonds|>
  count_wide(c(clarity, color), cut)
```

Plot Functions

```{r}
diamonds %>% 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

diamonds %>% 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.05)

histogram <- function(df, var, binwidth = NULL){
  df %>% 
    ggplot(aes(x = {{var}})) +
    geom_histogram(binwidth = binwidth)
}

diamonds %>% 
  histogram(carat, 0.1)

diamonds %>% 
  histogram(carat, 0.1) +
  labs(x = "Size in carats", y = "Number of diamonds")
```

More variables

```{r}
linearity_check <- function(df, x, y){
  df %>% 
    ggplot(aes(x = {{x}}, y = {{y}})) +
    geom_point() +
    geom_smooth(method = "loess", formula = y ~ x, color = "orange", se = FALSE) +
    geom_smooth(method = "lm", formula = y ~ x, color = "blue", se = FALSE)
}

starwars %>% 
  filter(mass < 1000) %>% 
  linearity_check(mass, height)

```

Cobining with other tidyverse

We have to use a new operator here, :=, because we are generating the variable name based on user-supplied data. Variable names go on the left hand side of =, but R’s syntax doesn’t allow anything to the left of = except for a single literal name. To work around this problem, we use the special operator := which tidy evaluation treats in exactly the same way as =.

```{r}
sorted_bars <- function(df, var){
  df %>% 
    mutate({{var}} := fct_rev(fct_infreq({{ var }})))} %>% 
  ggplot(aes(y = {{var}})) +
  geom_bar()

diamonds %>% 
  sorted_bars(clarity)
```

```{r}
conditional_bars <- function(df, condition, var){
  df %>% 
    filter({{condition}}) %>% 
    ggplot(aes(x = {{var}})) +
    geom_bar()
}

diamonds %>% 
  conditional_bars(cut == "Good", clarity)
```

Labeling

```{r}
histogram <- function(df, var, binwidth){
  label <- rlang::englue("A histogram of {{var}} with binwidth {binwidth}")
  
  df %>% 
    ggplot(aes(x = {{var}})) +
    geom_histogram(binwidth = binwidth) +
    labs(title = label)
}

diamonds %>% 
  histogram(carat, 0.1)
```

